---
title: "Supervised learning: deep learning"
date: last-modified
callout-appearance: simple
params:
  answers: false
format: 
  html: 
    toc: true
    toc-depth: 2
    self-contained: true
---

# Introduction

In this practical, we will create a feed-forward neural network as well as an optional convolutional neural network to analyze the famous MNIST dataset.

```{r}
#| label: packages
#| warning: false
#| message: false
library(tidyverse)
library(keras3)
```

::: callout-note
Let's set the seed value and use the same number as below to reproduce the same results.
:::

```{r}
#| label: seed
#| include: true
set.seed(45)
```

In this section, we will develop a deep feed-forward neural network for MNIST.

# Data preparation

::: {.callout-tip icon="false"}
## 1. Load the built-in MNIST dataset by running the following code. Then, describe the structure and contents of the `mnist` object.

```{r}
#| label: load_data
#| message: false

mnist <- dataset_mnist()
```

```{r}
#| label: q1
#| include: !expr params$answers

# str() function for inspecting the STRucture of an object
str(mnist)

# mnist is a list of 2 objects:
# - train (N=60000)
# - test (N=10000)

range(mnist$train$x)
table(mnist$train$y)

# in each object, there is an X with integers between 0-255 in a 28*28 matrix for each observation
# and a y which is a single integer (0-9) labelling each matrix
```
:::

:::: {.callout-tip icon="false"}
## 2. Plotting is very important when working with image data. We have defined a convenient plotting function for you. Use the `plot_img()` function below to plot the first training image. The `img` parameter has to be a matrix with dimensions `(28, 28)`.

::: callout-note
Indexing in 3-dimensional arrays works the same as indexing in matrices, but you need an extra comma `x[,,]`.
:::

```{r}
#| label: plot_img
#| message: false

plot_img <- function(img, col = gray.colors(255, start = 1, end = 0), ...) {
  image(t(img), asp = 1, ylim = c(1.1, -0.1), col = col, bty = "n", axes = FALSE, ...)
}
```

```{r}
#| label: q2
#| include: !expr params$answers

# get first training image (observation number is in first dimension)
im <- mnist$train$x[1,,]
plot_img(im)
```
::::

::: callout-note
It is usually a good idea to normalize your features to have a manageable, standard range before entering them in neural networks.
:::

::: {.callout-tip icon="false"}
## 3. As a preprocessing step, ensure the brightness values of the images in the training and test set are in the range (0, 1).

```{r}
#| label: q3
#| include: !expr params$answers

# remember that the range of training image values is 0 to 255
mnist$train$x <- mnist$train$x / 255
mnist$test$x <- mnist$test$x / 255

```
:::

# Multi-layer perceptron: multinomial logistic regression

The simplest neural network model is a multi-layer perceptron where we have no hidden layers and only input and output layers. We can call this a multinomial logistic regression model, where we have no hidden layers and 10 outputs (0-1) for our mnist data. That model is shown below.

```{r}
#| label: mlp
#| include: true

multinom  <- 
  # initialize a sequential model
  keras_model_sequential(input_shape = c(28, 28)) |> 
  # flatten 28*28 matrix into single vector
  layer_flatten() |>
  # softmax outcome == probability for each of 10 outputs
  layer_dense(10, activation = "softmax")

multinom$compile(
  loss = "sparse_categorical_crossentropy", # loss function for multinomial outcome
  optimizer = "adam", # we use this optimizer because it works well
  metrics = list("accuracy") # we want to know training accuracy in the end
)
```

::: {.callout-tip icon="false"}
## 4. Display a summary of the multinomial model using the `summary()` function. Describe why this model has 7850 parameters.

```{r}
#| label: q4
#| include: !expr params$answers

summary(multinom)

# parameters: 784*10 weights + 10 biases at output nodes = 7850

```
:::

::: {.callout-tip icon="false"}
## 5. Train the model for 5 epochs using the code below. What accuracy do we obtain in the validation set?

```{r}
#| label: fit_keras
#| output: false
#| eval: !expr params$answers

multinom |> 
  fit(x = mnist$train$x, 
      y = mnist$train$y, 
      epochs = 5, 
      validation_split = 0.2, 
      verbose = 1)

```

```{r}
#| label: q5
#| include: !expr params$answers

# about 92.69% accuracy on the validation set

```
:::

::: {.callout-tip icon="false"}
## 6. Train the model for another 5 epochs. What accuracy do we obtain in the validation set?

```{r}
#| label: q6
#| include: !expr params$answers
#| output: false
#| eval: !expr params$answers

multinom |> 
  fit(x = mnist$train$x, 
      y = mnist$train$y, 
      epochs = 5, 
      validation_split = 0.2, 
      verbose = 1)

# about 92.98% accuracy on the validation set. Slightly improved but not much!
```
:::

# Deep feed-forward neural networks

::: {.callout-tip icon="false"}
## 7. Create and compile a feed-forward neural network with the following properties. Ensure that the model has 50890 parameters.

-   sequential model
-   flatten layer
-   dense layer with 64 hidden units and "relu" activation function
-   dense output layer with 10 units and softmax activation function

You may reuse code from the multinomial model.

```{r}
#| label: q7
#| include: !expr params$answers

ffnn  <- 
  # initialize a sequential model
  keras_model_sequential(input_shape = c(28, 28)) |> 
  # flatten 28*28 matrix into single vector
  layer_flatten() |>
  # this is the hidden layer!
  layer_dense(64, activation = "relu") |>
  # softmax outcome == probability for each of 10 outputs
  layer_dense(10, activation = "softmax")

ffnn$compile(
  loss = "sparse_categorical_crossentropy", # loss function for multinomial outcome
  optimizer = "adam", # we use this optimizer because it works well
  metrics = list("accuracy") # we want to know training accuracy in the end
)

summary(ffnn)
```
:::

::: {.callout-tip icon="false"}
## 8. Train the model for 10 epochs. What do you see in terms of validation accuracy, also compared to the multinomial model?

```{r}
#| label: q8
#| include: !expr params$answers
#| output: false
#| eval: !expr params$answers

ffnn |> fit(x = mnist$train$x, y = mnist$train$y, epochs = 10, validation_split = 0.2, verbose = 1)
# in my case, higher accuracy! 97.12%
```
:::

::: {.callout-tip icon="false"}
## 9. Create predictions for the test data using the two trained models (using the function below). Create a confusion matrix and compute test accuracy for these two models. Write down any observations you have.

```{r}
#| label: class_predict
#| include: true

class_predict <- function(model, x_train) predict(model, x = x_train) |> apply(1, which.max) - 1

```

```{r}
#| label: q9
#| include: !expr params$answers

pred_multinom <- class_predict(multinom, x = mnist$test$x)
pred_ffnn     <- class_predict(ffnn, x = mnist$test$x)

(ctab_multinom <- table(pred = pred_multinom, true = mnist$test$y))
(ctab_ffnn     <- table(pred = pred_ffnn, true = mnist$test$y))

sum(diag(ctab_multinom)) / sum(ctab_multinom)
sum(diag(ctab_ffnn)) / sum(ctab_ffnn)

```
:::

::: {.callout-tip icon="false"}
## 10. Create and estimate (10 epochs) a deep feed-forward model with the following properties. Compare this model to the previous models on the test data.

-   sequential model
-   flatten layer
-   dense layer with 128 hidden units and "relu" activation function
-   dense layer with 64 hidden units and "relu" activation function
-   dense output layer with 10 units and softmax activation function

```{r}
#| label: q10
#| include: !expr params$answers
#| output: false
#| eval: !expr params$answers

dffnn  <- 
  keras_model_sequential(input_shape = c(28, 28)) |> # initialize a sequential model
  layer_flatten() |> # flatten 28*28 matrix into single vector
  layer_dense(128, activation = "relu") |> # this is the hidden layer!
  layer_dense(64, activation = "relu") |> # this is the hidden layer!
  layer_dense(10, activation = "softmax") # softmax outcome == logistic regression for each of 10 outputs

dffnn$compile(
  loss = "sparse_categorical_crossentropy", # loss function for multinomial outcome
  optimizer = "adam", # we use this optimizer because it works well
  metrics = list("accuracy") # we want to know training accuracy in the end
)

summary(dffnn)

dffnn |> fit(x = mnist$train$x, y = mnist$train$y, epochs = 10, validation_split = 0.2, verbose = 1)
```

```{r}
#| label: q10b
#| include: !expr params$answers
#| eval: !expr params$answers

pred_dffnn <- class_predict(dffnn, x = mnist$test$x)
(ctab_dffnn <- table(pred = pred_dffnn, true = mnist$test$y))
sum(diag(ctab_dffnn)) / sum(ctab_dffnn)

# this one does even better!
```
:::

# OPTIONAL: convolutional neural network

::: callout-note
Convolution layers in Keras need a specific form of data input.
:::

For each example, they need a `(width, height, channels)` array (tensor). For a colour image with 28\*28 dimension, that shape is usually `(28, 28, 3)`, where the channels indicate red, green, and blue. MNIST has no colour info, but we still need the channel dimension to enter the data into a convolution layer with shape `(28, 28, 1)`. The training dataset `x_train` should thus have shape `(60000, 28, 28, 1)`.

::: {.callout-tip icon="false"}
## 11. Add a "channel" dimension to the training and test data using the following code. Plot an image using the first channel of the 314th training example (this is a 9).

```{r}
#| label: add_channel

# add channel dimension to input (required for convolution layers)
dim(mnist$train$x) <- c(dim(mnist$train$x), 1)
dim(mnist$test$x)  <- c(dim(mnist$test$x), 1)

```

```{r}
#| label: q11
#| include: !expr params$answers

plot_img(mnist$train$x[314,,,1])
```
:::

::: {.callout-tip icon="false"}
## 12. Create and compile a convolutional neural network using the following code. Describe the different layers in your own words.

```{r}
#| label: cnn

cnn <- 
  keras_model_sequential(input_shape = c(28, 28, 1)) |> 
  layer_conv_2d(filters = 6, kernel_size = c(5, 5)) |> 
  layer_max_pooling_2d(pool_size = c(4, 4)) |>
  layer_flatten() |> 
  layer_dense(units = 32, activation = "relu") |> 
  layer_dense(10, activation = "softmax")

cnn |> 
  compile(
    loss = "sparse_categorical_crossentropy",
    optimizer = "adam", 
    metrics = c("accuracy")
  )

```

```{r}
#| label: q12
#| include: !expr params$answers

summary(cnn)

# First, we have the input layer which gets the images and the first channel (28, 28, 1)
# then, there is a 2d convolution layer with 6 filters, and a kernel size of 5 (in each direction)
# then, we max-pool the resulting 6 maps to reduce their size by 4 in each direction
# afterwards, we flatten
# then comes a dense hidden layer with 32 units and a relu activation function
# lastly, the output layer is the same as before

```
:::

::: {.callout-tip icon="false"}
## 13. Fit this model on the training data (10 epochs) and compare it to the previous models.

```{r}
#| label: q13
#| include: !expr params$answers
#| output: false
#| eval: !expr params$answers

cnn |> fit(x = mnist$train$x, y = mnist$train$y, epochs = 10, validation_split = 0.2, verbose = 1)
```

```{r}
#| label: q13b
#| include: !expr params$answers

pred_cnn <- class_predict(cnn, x = mnist$test$x)
(ctab_cnn <- table(pred = pred_cnn, true = mnist$test$y))
sum(diag(ctab_cnn)) / sum(ctab_cnn)

# has about the same performance as the deep feed-forward model (mind you - with way fewer parameters!)
```
:::

::: {.callout-tip icon="false"}
## 14. Create another CNN which has better performance within 10 epochs. Compare your validation or test accuracy to that of your peers.

Here are some things you could do:

-   Reduce the convolution filter size & the pooling size and add a second convolutional & pooling layer with double the number of filters
-   Add a dropout layer after the flatten layer
-   Look up on the internet what works well and implement it!

```{r}
#| label: q14
#| include: !expr params$answers
#| output: false
#| eval: !expr params$answers

cnn_2 <- 
  keras_model_sequential(input_shape = c(28, 28, 1)) |> 
  layer_conv_2d(filters = 6, kernel_size = c(3, 3)) |> 
  layer_max_pooling_2d(pool_size = c(2, 2)) |>
  layer_conv_2d(filters = 12, kernel_size = c(3, 3)) |> 
  layer_max_pooling_2d(pool_size = c(2, 2)) |>
  layer_flatten() |> 
  layer_dropout(rate = 0.2) |> 
  layer_dense(units = 32, activation = "relu") |> 
  layer_dense(10, activation = "softmax")

cnn_2 |> 
  compile(
    loss = "sparse_categorical_crossentropy",
    optimizer = "adam", 
    metrics = c("accuracy")
  )

cnn_2 |> fit(x = mnist$train$x, y = mnist$train$y, epochs = 10, validation_split = 0.2, verbose = 1)
```

```{r}
#| label: q14b
#| include: !expr params$answers
#| eval: !expr params$answers

pred_cnn_2 <- class_predict(cnn_2, x = mnist$test$x)
(ctab_cnn_2 <- table(pred = pred_cnn_2, true = mnist$test$y))
sum(diag(ctab_cnn_2)) / sum(ctab_cnn_2)

```
:::
