---
title: "Hierarchical and k-means clustering"
date: last-modified
callout-appearance: simple
params:
  answers: true
format: 
  html: 
    toc: true
    toc-depth: 2
    self-contained: true
---

# Introduction

In this practical, we will apply hierarchical and k-means clustering to two synthetic datasets. 

First, we load the packages required for this lab.

```{r}
#| label: packages
#| warning: false
#| message: false
library(MASS)
library(tidyverse)
library(patchwork)
library(ggdendro)
```

::: {.callout-note}
Make sure to load `MASS` before `tidyverse` otherwise the function `MASS::select()` will overwrite `dplyr::select()` 
:::

Before we start, set a seed for reproducibility, we use `123`, and also use `options(scipen = 999)` to suppress scientific notations, making it easier to compare and interpret results later in the session.

```{r}
#| label: seed
set.seed(123)
options(scipen = 999)
```

# Data processing

::: {.callout-tip icon=false}
## 1. The data can be generated by running the code below. Try to understand what is happening as you run each line of the code below.

```{r}
#| label: q1

# randomly generate bivariate normal data
set.seed(123)
sigma      <- matrix(c(1, .5, .5, 1), 2, 2)
sim_matrix <- mvrnorm(n = 100, mu = c(5, 5), Sigma = sigma)
colnames(sim_matrix) <- c("x1", "x2")

# change to a data frame (tibble) and add a cluster label column
sim_df <- 
  sim_matrix |> 
  as_tibble() |>
  mutate(class = sample(c("A", "B", "C"), size = 100, replace = TRUE))

# Move the clusters to generate separation
sim_df_small <- 
  sim_df |>
  mutate(x2 = case_when(class == "A" ~ x2 + .5,
                        class == "B" ~ x2 - .5,
                        class == "C" ~ x2 + .5),
         x1 = case_when(class == "A" ~ x1 - .5,
                        class == "B" ~ x1 - 0,
                        class == "C" ~ x1 + .5))
sim_df_large <- 
  sim_df |>
  mutate(x2 = case_when(class == "A" ~ x2 + 2.5,
                        class == "B" ~ x2 - 2.5,
                        class == "C" ~ x2 + 2.5),
         x1 = case_when(class == "A" ~ x1 - 2.5,
                        class == "B" ~ x1 - 0,
                        class == "C" ~ x1 + 2.5))
```
:::

::: {.callout-tip icon=false}

## 2. Prepare two unsupervised datasets by removing the class feature.

```{r}
#| label: q2
#| include: !expr params$answers
df_s <- sim_df_small |> select(-class)
df_l <- sim_df_large |> select(-class)
```

:::

::: {.callout-tip icon=false}
## 3. For each of these datasets, create a scatterplot. Combine the two plots into a single frame (look up the "patchwork" package to see how to do this!) What is the difference between the two datasets?

```{r}
#| label: q3
#| include: !expr params$answers
# patchwork defines the "+" operator to combine entire ggplots!
df_s |> ggplot(aes(x = x1, y = x2)) + geom_point() + ggtitle("Small") +
df_l |> ggplot(aes(x = x1, y = x2)) + geom_point() + ggtitle("Large")

# df_s has a lot of class overlap, df_l has very little overlap
```
:::


# Hierarchical clustering

::: {.callout-tip icon=false}
## 4. Run a hierarchical clustering on these datasets and display the result as dendrograms. Use euclidian distances and the complete agglomeration method. Make sure the two plots have the same y-scale. What is the difference between the dendrograms?

Hint: functions you will need are `hclust`, `ggdendrogram`, and `ylim`.

```{r}
#| label: q4
#| include: !expr params$answers
dist_s <- dist(df_s, method = "euclidian")
dist_l <- dist(df_l, method = "euclidian")

res_s <- hclust(dist_s, method = "complete")
res_l <- hclust(dist_l, method = "complete")

ggdendrogram(res_s, labels = FALSE) + ggtitle("Small") + ylim(0, 10) +
ggdendrogram(res_l, labels = FALSE) + ggtitle("Large") + ylim(0, 10)

# The dataset with large differences segments into 3 classes much higher up.
# Interestingly, the microstructure (lower splits) is almost exactly the same,
# because within the three clusters there is no difference between the datasets.
```
:::

::: {.callout-tip icon=false}
## 5. For the dataset with small differences, also run a complete agglomeration hierarchical cluster with manhattan distance.

```{r}
#| label: q5
#| include: !expr params$answers
dist_s2 <- dist(df_s, method = "manhattan")
res_s2  <- hclust(dist_s2, method = "complete")
```
:::

::: {.callout-tip icon=false}
## 6. Use the `cutree()` function to obtain the cluster assignments for three clusters and compare the cluster assignments to the 3-cluster euclidian solution. Do this comparison by creating two scatter plots with cluster assignment mapped to the colour aesthetic. Which difference do you see?

```{r}
#| label: q6
#| include: !expr params$answers
clus_1 <- as_factor(cutree(res_s, 3))
clus_2 <- as_factor(cutree(res_s2, 3))

p1 <- df_s |> 
  ggplot(aes(x = x1, y = x2, colour = clus_1)) + 
  geom_point() + 
  ggtitle("Euclidian") + 
  theme(legend.position = "n") +
  coord_fixed()

p2 <- df_s |> 
  ggplot(aes(x = x1, y = x2, colour = clus_2)) + 
  geom_point() + 
  ggtitle("Manhattan") + 
  theme(legend.position = "n") +
  coord_fixed()

p1 + p2

# The manhattan distance clustering prefers more rectangular classes, whereas
# the euclidian distance clustering prefers circular classes. The difference is
# most prominent in the very center of the plot and for the top right cluster.
```
:::

# K-means clustering

::: {.callout-tip icon=false}
## 7. Create k-means clustering with 2, 3, 4, and 6 clusters on the large difference data. Again, create coloured scatter plots for these clustering results.

```{r}
#| label: q7
#| include: !expr params$answers
# Let's set the seed for reproducibility
set.seed(45)
# we can do it in a single pipeline and with faceting 
# (of course there are many ways to do this, though)
df_l |> 
  mutate(
    class_2 = as_factor(kmeans(df_l, 2)$cluster),
    class_3 = as_factor(kmeans(df_l, 3)$cluster),
    class_4 = as_factor(kmeans(df_l, 4)$cluster),
    class_6 = as_factor(kmeans(df_l, 6)$cluster)
  ) |> 
  pivot_longer(cols = c(class_2, class_3, class_4, class_6), 
               names_to = "class_num", values_to = "cluster") |> 
  ggplot(aes(x = x1, y = x2, colour = cluster)) +
  geom_point() +
  scale_colour_brewer(type = "qual") + # use easy to distinguish scale
  facet_wrap(~class_num)
```
:::

::: {.callout-tip icon=false}
## 8. Do the same thing again a few times. Do you see the same results every time? where do you see differences?

```{r}
#| label: q8
#| include: !expr params$answers
# Let's set the seed for reproducibility
set.seed(46)
df_l |> 
  mutate(
    class_2 = as_factor(kmeans(df_l, 2)$cluster),
    class_3 = as_factor(kmeans(df_l, 3)$cluster),
    class_4 = as_factor(kmeans(df_l, 4)$cluster),
    class_6 = as_factor(kmeans(df_l, 6)$cluster)
  ) |> 
  pivot_longer(cols = c(class_2, class_3, class_4, class_6), 
               names_to = "class_num", values_to = "cluster") |> 
  ggplot(aes(x = x1, y = x2, colour = cluster)) +
  geom_point() +
  scale_colour_brewer(type = "qual") + # use easy to distinguish scale
  facet_wrap(~class_num)

# There is label switching in all plots. There is a different result altogether
# in the class_4 solution in this case.
```
:::

::: {.callout-tip icon=false}
## 9. Find a way online to perform bootstrap stability assessment for the 3 and 6-cluster solutions.

```{r}
#| label: q9
#| include: !expr params$answers
# For this we decided to use the function clusterboot from the fpc package.
# NB: this package needs to be installed first!
# install.packages("fpc")
library(fpc)

# You can read the documentation ?clusterboot to figure out how to use this 
# function. This can take a while but is something you will need to do a lot in
# the real world!
boot_3 <- clusterboot(df_l, B = 1000, clustermethod = kmeansCBI, k = 3, 
                      count = FALSE)
boot_6 <- clusterboot(df_l, B = 1000, clustermethod = kmeansCBI, k = 6, 
                      count = FALSE)

# the average stability is much lower for 6 means than for 3 means:
boot_3$bootmean
boot_6$bootmean
```
:::
