---
title: "Data visualization using ggplot2"
date: last-modified
callout-appearance: simple
params:
  answers: true
format: 
  html: 
    toc: true
    toc-depth: 2
    self-contained: true
---

## Introduction

In this practical, we will learn how to create data visualisations using the grammar of graphics. For the visualisations, we will be using a package that implements the grammar of graphics: `ggplot2`. This package is part of the `tidyverse`, so we only need to load `tidyverse`. We will use datasets from the `ISLR` package.

```{r}
#| label: packages
#| warning: false
#| message: false
library(ISLR)
library(tidyverse)
```


::: {.callout-note}
An excellent reference manual for `ggplot` can be found on the tidyverse website: [https://ggplot2.tidyverse.org/reference/](https://ggplot2.tidyverse.org/reference/)
:::

## What is `ggplot`?

Plots can be made in `R` without the use of `ggplot` using `plot()`, `hist()` or `barplot()` and related functions. Here is an example of each on the `Hitters` dataset from `ISLR`:

```{r}
#| label: baseplots

# Get an idea of what the Hitters dataset looks like
head(Hitters)

# histogram of the distribution of salary
hist(Hitters$Salary, xlab = "Salary in thousands of dollars")

# barplot of how many members in each league
barplot(table(Hitters$League))


# Number of career hits versus number of career home runs
plot(x = Hitters$Hits, y = Hitters$HmRun, 
     xlab = "Hits", ylab = "Home runs")
```

These plots are informative and useful for visually inspecting the dataset, and they each have a specific syntax associated with them. `ggplot` has a more unified approach to plotting, where you build up a plot layer by layer using the `+` operator:

```{r}
#| label: ggplot example
homeruns_plot <- 
  ggplot(Hitters, aes(x = Hits, y = HmRun)) +
  geom_point() +
  labs(x = "Hits", y = "Home runs")

homeruns_plot
```

As introduced in the lecture, a `ggplot` object is built up in different layers:

1. input the dataset to a `ggplot()` function call
2. construct aesthetic mappings
3. add (geometric) components to your plot that use these mappings
4. add labels, themes, visuals.

Because of this layered syntax, it is then easy to add elements like these fancy density lines, a title, and a different theme:

```{r}
#| label: ggplot example 2
homeruns_plot + 
  geom_density_2d() +
  labs(title = "Cool density and scatter plot of baseball data") +
  theme_minimal()
```

In conclusion, `ggplot` objects are easy to manipulate and they force a principled approach to data visualisation. In this practical, we will learn how to construct them.

::: {.callout-tip icon=false}

## 1. Name the aesthetics, geoms, scales, and facets of the above visualisation. Also name any statistical transformations or special coordinate systems.

```{r}
#| label: q1
#| include: !expr params$answers

# Aesthetics
#   number of hits mapped to x-position
#   number of home runs mapped to y-position
# Geoms: points and contour lines
# Scales:
#   x-axis: continuous
#   y-axis: continuous
# Facets: None
# Statistical transformations: None
# Special coordinate system: None (just cartesian)
```

:::

## Aesthetics and data preparation

The first step in constructing a `ggplot` is the preparation of your data and the mapping of variables to aesthetics. In the `homeruns_plot`, we used an existing data frame, the `Hitters` dataset. 

The data frame needs to have proper column names and the types of the variables in the data frame need to be correctly specified. Numbers should be `numerics`, categories should be `factors`, and names or identifiers should be `character` variables. `ggplot()` _always_ expects a data frame, which may feel awfully strict, but it allows for excellent flexibility in the remaining plotting steps.

::: {.callout-tip icon=false}

## 2. Run the code below to generate data. 

```{r}
#| label: datavecs
set.seed(1234)
student_grade  <- rnorm(32, 7)
student_number <- round(runif(32) * 2e6 + 5e6)
programme      <- sample(c("Science", "Social Science"), 32, replace = TRUE)
```

There will be three vectors in your environment. Put them in a data frame for entering it in a `ggplot()` call using either the `data.frame()` or the `tibble()` function. Give informative names and make sure the types are correct (use the `as.<type>()` functions). Name the result `gg_students`.

```{r}
#| label: prepare
#| include: !expr params$answers
gg_students <- tibble(
  number = as.character(student_number), # an identifier
  grade  = student_grade,                # already the correct type.
  prog   = as.factor(programme)          # categories should be factors.
)

head(gg_students)

# note that if you use data.frame(), you need to set the argument 
# stringsAsFactors to FALSE to get student number to be a character.
# tibble() does this by default.
```

(we will use this dataset later in the practical)

:::


__Mapping aesthetics__ is usually done in the main `ggplot()` call. Aesthetic mappings are the second argument to the function, after the data. 


::: {.callout-tip icon=false}

## 3. Plot the first `homeruns_plot` again, but map the `Hits` to the y-axis and the `HmRun` to the x-axis instead.

```{r}
#| label: hmrn2
#| include: !expr params$answers
ggplot(Hitters, aes(x = HmRun, y = Hits)) +
  geom_point() +
  labs(y = "Hits", x = "Home runs")
```

:::



::: {.callout-tip icon=false}

## 4. Recreate the same plot once more, but now also map the variable `League` to the `colour` aesthetic and the variable `Salary` to the `size` aesthetic.

```{r}
#| label: hmrn3
#| include: !expr params$answers
ggplot(Hitters, aes(x = HmRun, y = Hits, colour = League, size = Salary)) +
  geom_point() +
  labs(y = "Hits", x = "Home runs")
```

:::


Examples of aesthetics are:

- x
- y
- alpha (transparency)
- colour
- fill
- group
- shape
- size
- stroke


## Geoms

Up until now we have used two geoms: contour lines and points. The geoms in `ggplot2` are added via the `geom_<geomtype>()` functions. Each geom has a required aesthetic mapping to work. For example, `geom_point()` needs at least and x and y position mapping, as you can read [here](https://ggplot2.tidyverse.org/reference/geom_point.html#aesthetics). You can check the required aesthetic mapping for each geom via the `?geom_<geomtype>`.


::: {.callout-tip icon=false}

## 5. Look at the many different geoms on the [reference website](https://ggplot2.tidyverse.org/reference/#section-layer-geoms).

:::

There are two types of geoms: 

- geoms which perform a transformation of the data beforehand, such as `geom_density_2d()` which calculates contour lines from x and y positions.
- geoms which do not transform data beforehand, but use the aesthetic mapping directly, such as `geom_point()`.

## Histogram


::: {.callout-tip icon=false}

## 6. Use `geom_histogram()` to create a histogram of the grades of the students in the `gg_students` dataset. 

Play around with the `binwidth` argument of the `geom_histogram()` function.

```{r}
#| label: geomhist
#| include: !expr params$answers
gg_students |>
  ggplot(aes(x = grade)) +
  geom_histogram(binwidth = .5)
```

:::





## Density


The continuous equivalent of the histogram is the density estimate.

::: {.callout-tip icon=false}

## 7. Use `geom_density()` to create a density plot of the grades of the students in the `gg_students` dataset. Add the argument `fill = "light seagreen"` to `geom_density()`.


```{r}
#| label: geomdens
#| include: !expr params$answers
gg_students |> 
  ggplot(aes(x = grade)) +
  geom_density(fill = "light seagreen")
```

:::

The downside of only looking at the density or histogram is that it is an abstraction from the raw data, thus it might alter interpretations. For example, it could be that a grade between 8.5 and 9 is in fact impossible. We do not see this in the density estimate. To counter this, we can add a raw data display in the form of rug marks.

::: {.callout-tip icon=false}

## 8. Add rug marks to the density plot through `geom_rug()`. You can edit the `colour` and `size` of the rug marks using those arguments within the `geom_rug()` function.


```{r}
#| label: geomdens2
#| include: !expr params$answers
gg_students |> 
  ggplot(aes(x = grade)) +
  geom_density(fill = "light seagreen") +
  geom_rug(size = 1, colour = "light seagreen")
```

:::

::: {.callout-tip icon=false}

## 9. Increase the data to ink ratio by removing the y axis label, setting the theme to `theme_minimal()`, and removing the border of the density polygon.

Also set the limits of the x-axis to go from 0 to 10 using the `xlim()` function, because those are the plausible values for a student grade.

```{r}
#| label: geomdens3
#| include: !expr params$answers
gg_students |> 
  ggplot(aes(x = grade)) +
  geom_density(fill = "light seagreen", colour = NA) +
  geom_rug(size = 1, colour = "light seagreen") +
  theme_minimal() +
  labs(y = "") +
  xlim(0, 10)
```

:::


## Boxplot

A common task is to compare distributions across groups. A classic example of a visualisation that performs this is the boxplot, accessible via `geom_boxplot()`. It allows for visual comparison of the distribution of two or more groups through their summary statistics.


::: {.callout-tip icon=false}

## 10. Create a boxplot of student grades per programme in the `gg_students` dataset you made earlier: map the programme variable to the x position and the grade to the y position. 

For extra visual aid, you can additionally map the programme variable to the fill aesthetic.

```{r}
#| label: boxplot
#| include: !expr params$answers
gg_students |> 
  ggplot(aes(x = prog, y = grade, fill = prog)) +
  geom_boxplot() +
  theme_minimal()
```

:::


::: {.callout-tip icon=false}

## 11. What do each of the horizontal lines in the boxplot mean? What do the vertical lines (whiskers) mean?


```{r}
#| label: bplot
#| include: !expr params$answers
# From the help file of geom_boxplot:

# The middle line indicates the median, the outer horizontal
# lines are the 25th and 75th percentile.

# The upper whisker extends from the hinge to the largest value 
# no further than 1.5 * IQR from the hinge (where IQR is the 
# inter-quartile range, or distance between the first and third 
# quartiles). The lower whisker extends from the hinge to the 
# smallest value at most 1.5 * IQR of the hinge. Data beyond 
# the end of the whiskers are called "outlying" points and are 
# plotted individually.
```

:::

## Bar plot

We can display amounts or proportions as a bar plot to compare group sizes of a factor.


::: {.callout-tip icon=false}

## 12. Create a bar plot of the variable `Years` from the `Hitters` dataset.

```{r}
#| label: ggbarplot
#| include: !expr params$answers
Hitters |> 
  ggplot(aes(x = Years)) + 
  geom_bar() +
  theme_minimal()
```

:::

`geom_bar()` automatically transforms variables to counts (see `?stat_count`), similar to how the function `table()` works:

```{r}
#| label: table
table(Hitters$Years)
```

# Faceting


::: {.callout-tip icon=false}

## 13. Create a data frame called `baseball` based on the `Hitters` dataset. 

In this data frame, create a factor variable which splits players' salary range into 3 categories. Tip: use the `filter()` function to remove the missing values, and then use the `cut()` function and assign nice `labels` to the categories. In addition, create a variable which indicates the proportion of career hits that was a home run.

```{r}
#| label: cut
#| include: !expr params$answers
baseball <-
  Hitters |> 
  filter(!is.na(Salary)) |> 
  mutate(
    Salary_range = cut(Salary, breaks = 3, 
                       labels = c("Low salary", "Mid salary", "High salary")),
    Career_hmrun_proportion = CHmRun/CHits
  )
```

:::

::: {.callout-tip icon=false}

## 14. Create a scatter plot where you map `CWalks` to the x position and the proportion you calculated in the previous exercise to the y position. 

Fix the y axis limits to (0, 0.4) and the x axis to (0, 1600) using `ylim()` and `xlim()`. Add nice x and y axis titles using the `labs()` function. Save the plot as the variable `baseball_plot`.


```{r}
#| label: scatter
#| include: !expr params$answers
baseball_plot <-   
  baseball |> 
  ggplot(aes(x = CWalks, y = Career_hmrun_proportion)) +
  geom_point() +
  ylim(0, 0.4) +
  xlim(0, 1600) + 
  theme_minimal() +
  labs(y = "Proportion of home runs",
       x = "Career number of walks")

baseball_plot
```

:::


::: {.callout-tip icon=false}

## 15. Split up this plot into three parts based on the salary range variable you calculated. Use the `facet_wrap()` function for this; look at the examples in the help file for tips.

```{r}
#| label: facet
#| include: !expr params$answers
baseball_plot + facet_wrap(~Salary_range)
```

:::

Faceting can help interpretation. In this case, we can see that high-salary earners are far away from the point (0, 0) on average, but that there are low-salary earners which are even further away. Faceting should preferably be done using a `factor` variable. The order of the facets is taken from the `levels()` of the factor. Changing the order of the facets can be done using `fct_relevel()` if needed.

## OPTIONAL: Final exercise

If you have time left, do this final exercise!

::: {.callout-tip icon=false}

## 16. Create an interesting data visualisation based on the `Carseats` data from the `ISLR` package.

```{r}
#| label: final
#| include: !expr params$answers
# an example answer could be:
Carseats |> 
  mutate(Competition = Price/CompPrice,
         ShelveLoc   = fct_relevel(ShelveLoc, "Bad", "Medium", "Good")) |> 
  ggplot(aes(x = Competition, y = Sales, colour = Age)) +
  geom_point() +
  theme_minimal() +
  scale_colour_viridis_c() + # add a custom colour scale
  facet_wrap(vars(ShelveLoc))
```

:::
