---
title: "Model-based clustering using mclust"
date: last-modified
callout-appearance: simple
params:
  answers: true
format: 
  html: 
    toc: true
    toc-depth: 2
    self-contained: true
---

# Introduction

In this practical, we will apply model-based clustering on a data set of bank note measurements.

We use the following packages:

```{r}
#| label: packages
#| warning: false
#| message: false
library(mclust)
library(tidyverse)
library(patchwork)
```


The data is built into the `mclust` package and can be loaded as a `tibble` by running the following code:


```{r}
#| label: data
df <- as_tibble(banknote)
```


# Data exploration

::: {.callout-tip icon=false}
## 1. Read the help file of the `banknote` data set to understand what it's all about.

```{r}
#| label: q1
#| include: !expr params$answers
#| eval: false
?banknote
```
:::

::: {.callout-tip icon=false}

## 2. Create a scatter plot of the left (x-axis) and right (y-axis) measurements on the data set. Map the `Status` column to colour. Jitter the points to avoid overplotting. Are the classes easy to distinguish based on these features?

```{r}
#| label: q2
#| include: !expr params$answers
df |> 
  ggplot(aes(x = Left, y = Right, colour = Status)) +
  geom_point(position = position_jitter())

# the classes are not easy to distinguish: there is considerable overlap.
```

:::

::: {.callout-tip icon=false}
## 3. From now on, we will assume that we don't have the labels. Remove the `Status` column from the data set.

```{r}
#| label: q3
#| include: !expr params$answers
df <- df |> select(-Status)
```
:::


::: {.callout-tip icon=false}
## 4. Create density plots for all columns in the data set. Which single feature is likely to be best for clustering?

```{r}
#| label: q4
#| include: !expr params$answers
# big patchwork of density plots
df |> ggplot(aes(x = Length))   + geom_density() + theme_minimal() +
df |> ggplot(aes(x = Left))     + geom_density() + theme_minimal() +
df |> ggplot(aes(x = Right))    + geom_density() + theme_minimal() +
df |> ggplot(aes(x = Bottom))   + geom_density() + theme_minimal() +
df |> ggplot(aes(x = Top))      + geom_density() + theme_minimal() +
df |> ggplot(aes(x = Diagonal)) + geom_density() + theme_minimal()
  
# the Diagonal feature looks good! Look at the two bumps in its density plot.

# Colourful alternative:
library(ggridges)
df |> 
  mutate(across(everything(), scale)) |> 
  pivot_longer(everything(), names_to = "feature", values_to = "value") |> 
  ggplot(aes(x = value, y = feature, fill = feature)) + 
  geom_density_ridges() +
  scale_fill_viridis_d(guide = "none") +
  theme_minimal()
```
:::

# Univariate model-based clustering

::: {.callout-tip icon=false}
## 5. Use `Mclust` to perform model-based clustering with 2 clusters on the feature you chose. Assume equal variances. Name the model object `fit_E_2`. What are the means and variances of the clusters?

```{r}
#| label: q5
#| include: !expr params$answers
fit_E_2 <- Mclust(df$Diagonal, G = 2, modelNames = "E")
# means:
fit_E_2$parameters$mean
# variances:
fit_E_2$parameters$variance$sigmasq
```
:::

::: {.callout-tip icon=false}
## 6. Use the formula from the slides and the model's log-likelihood (`fit_E_2$loglik`) to compute the BIC for this model. Compare it to the BIC stored in the model object (`fit_E_2$bic`). Explain how many parameters (m) you used and which parameters these are.

```{r}
#| label: q6
#| include: !expr params$answers
# This model has 4 parameters: 1 class probability pi, 2 means, and 1 variance
- 2 * fit_E_2$loglik + log(nrow(df)) * 4 
fit_E_2$bic

# note: the BIC from the Mclust package is (wrongly) multiplied by -1!
```
:::

::: {.callout-tip icon=false}
## 7. Plot the model-implied density using the `plot()` function. Afterwards, add rug marks of the original data to the plot using the `rug()` function from the base graphics system.

```{r}
#| label: q7
#| include: !expr params$answers
# plot the density using base R plots
plot(fit_E_2, "density", xlab = "Diagonal measurement")
# add the observations using rug marks
rug(df$Diagonal)
```
:::

::: {.callout-tip icon=false}
## 8. Use `Mclust` to perform model-based clustering with 2 clusters on this feature again, but now assume _unequal_ variances. Name the model object `fit_V_2`. What are the means and variances of the clusters? Plot the density again and note the differences.

```{r}
#| label: q8
#| include: !expr params$answers
# Let's set the seed for reproducibility
fit_V_2 <- Mclust(df$Diagonal, G = 2, modelNames = "V")
fit_V_2$parameters$mean
fit_V_2$parameters$variance$sigmasq
plot(fit_V_2, "density", xlab = "Diagonal measurement")
rug(df$Diagonal)
# The left cluster has larger variance now, because the data is more spread out
```
:::

::: {.callout-tip icon=false}
## 9. How many parameters does this model have? Name them.

```{r}
#| label: q9
#| include: !expr params$answers
# 5 parameters:
# 1 class probability (pi)
# 2 means
# 2 variances
```
:::

::: {.callout-tip icon=false}
## 10. According to the deviance, which model fits better?

```{r}
#| label: q10
#| include: !expr params$answers
-2*fit_E_2$loglik
-2*fit_V_2$loglik
# The V model fits better (lower deviance)
```
:::

::: {.callout-tip icon=false}
## 11. According to the BIC, which model is better?

```{r}
#| label: q11
#| include: !expr params$answers
# remember, the package shows negative BIC
-fit_E_2$bic
-fit_V_2$bic
# the V model fits better (lower BIC), but it's closer because of the extra parameter!
```
:::

# Multivariate model-based clustering

We will now use all available information in the data set to cluster the observations. 

::: {.callout-tip icon=false}
## 12. Use Mclust with all 6 features to perform clustering. Allow all model types (shapes), and from 1 to 9 potential clusters. What is the optimal model based on the BIC?

```{r}
#| label: q12
#| include: !expr params$answers
fit <- Mclust(df)
summary(fit)
# The optimal model is a VVE model with 3 components
```
:::

::: {.callout-tip icon=false}
## 13. How many mean parameters does this model have?

```{r}
#| label: q13
#| include: !expr params$answers
fit$parameters$mean
# 3 clusters * 6 variables = 18 mean parameters
```
:::

::: {.callout-tip icon=false}
## 14. Run a 2-component VVV model on this data. Create a matrix of bivariate contour ("density") plots using the `plot()` function. Which features provide good component separation? Which do not?

```{r}
#| label: q14
#| include: !expr params$answers
VVV2 <- Mclust(df, 2, "VVV")
plot(VVV2, "density")
# Diagonal-top, and diagonal-bottom separate the classes very well!
```
:::


::: {.callout-tip icon=false}
## 15. Create a scatter plot just like the first scatter plot in this tutorial, but map the estimated class assignments to the colour aesthetic. 

Map the uncertainty (part of the fitted model list) to the size aesthetic, such that larger points indicate more uncertain class assignments. Jitter the points to avoid overplotting. What do you notice about the uncertainty?

```{r}
#| label: q15
#| include: !expr params$answers
df |> 
  ggplot(aes(
    x      = Left, 
    y      = Right,
    colour = as_factor(VVV2$classification), 
    size   = VVV2$uncertainty 
  )) +
  geom_point(position = position_jitter()) +
  scale_size(range = c(1, 3)) +
  labs(colour = "Cluster", size = "Uncertainty")

# only 3 points are slightly uncertain, and they are not 
# around the border of the classes in these two dimensions. 
# Apparently, the other dimensions give enough information 
# about those points near the border!
```
:::