---
title: "Supervised learning: classification"
date: last-modified
callout-appearance: simple
params:
  answers: true
format: 
  html: 
    toc: true
    toc-depth: 2
    self-contained: true
---

# Introduction

In this practical, we will learn to use three different classification methods: K-nearest neighbours, logistic regression, and linear discriminant analysis.

One of the packages we are going to use is `class`. For this, you will probably need to `install.packages("class")` before running the `library()` functions.

```{r}
#| label: packages
#| warning: false
#| message: false
library(MASS)
library(class)
library(ISLR)
library(tidyverse)
```

::: {.callout-note}
Make sure to load `MASS` before `tidyverse` otherwise the function `MASS::select()` will overwrite `dplyr::select()` 
:::

```{r}
#| label: seed
#| include: false
set.seed(45)
```

# Default dataset

The default dataset contains credit card loan data for 10 000 people. The goal is to classify credit card cases as `yes` or `no` based on whether they will default on their loan.

::: {.callout-tip icon=false}
## 1. Create a scatterplot of the `Default` dataset, where `balance` is mapped to the x position, `income` is mapped to the y position, and `default` is mapped to the colour. Can you see any interesting patterns already?

```{r}
#| label: q1
#| include: !expr params$answers

Default |>  
  arrange(default) |> # so the yellow dots are plotted after the blue ones
  ggplot(aes(x = balance, y = income, colour = default)) +
  geom_point(size = 1.3) +
  theme_minimal() +
  scale_colour_viridis_d() # optional custom colour scale

# People with high remaining balance are more likely to default. 
# There seems to be a low-income group and a high-income group
```
:::


::: {.callout-tip icon=false}

## 2. Add `facet_grid(cols = vars(student))` to the plot. What do you see?

```{r}
#| label: q2
#| include: !expr params$answers

Default |> 
  arrange(default) |> # so the yellow dots are plotted after the blue ones
  ggplot(aes(x = balance, y = income, colour = default)) +
  geom_point(size = 1.3) +
  theme_minimal() +
  scale_colour_viridis_d() +
  facet_grid(cols = vars(student))

# The low-income group is students!
```

:::

::: {.callout-tip icon=false}
## 3. Transform "student" into a dummy variable using `ifelse()` (0 = not a student, 1 = student). Then, randomly split the Default dataset into a training set `default_train` (80%) and a test set `default_test` (20%)

```{r}
#| label: q3
#| include: !expr params$answers

default_df <- 
  Default |> 
  mutate(student = ifelse(student == "Yes", 1, 0)) |> 
  mutate(split = sample(rep(c("train", "test"), times = c(8000, 2000))))

default_train <- 
  default_df |> 
  filter(split == "train") |> 
  select(-split)

default_test <- 
  default_df |> 
  filter(split == "test") |> 
  select(-split)

```

:::


# K-Nearest Neighbours

Now that we have explored the dataset, we can start on the task of classification. We can imagine a credit card company wanting to predict whether a customer will default on the loan so they can take steps to prevent this from happening.

The first method we will be using is k-nearest neighbours (KNN). It classifies datapoints based on a majority vote of the k points closest to it. In `R`, the `class` package contains a `knn()` function to perform knn.

::: {.callout-tip icon=false}
## 4. Create class predictions for the test set using the `knn()` function. 

Use `student`, `balance`, and `income` (but no functions of those variables) in the `default_train` dataset. Set k to 5. Store the predictions in a variable called `knn_5_pred`.

```{r}
#| label: q4
#| include: !expr params$answers

knn_5_pred <- knn(
  train = default_train |> select(-default),
  test  = default_test  |> select(-default),
  cl    = as_factor(default_train$default),
  k     = 5
)
```

:::

::: {.callout-tip icon=false}
## 5. Create two scatter plots with income and balance as in the first plot you made. One with the true class (`default`) mapped to the colour aesthetic, and one with the predicted class (`knn_5_pred`) mapped to the colour aesthetic. 

Hint: Add the predicted class `knn_5_pred` to the `default_test` dataset before starting your `ggplot()` call of the second plot. What do you see?

```{r}
#| label: q5
#| include: !expr params$answers

# first plot is the same as before
default_test |> 
  arrange(default) |> 
  ggplot(aes(x = balance, y = income, colour = default)) +
  geom_point(size = 1.3) + 
  scale_colour_viridis_d() +
  theme_minimal() +
  labs(title = "True class")

# second plot maps pred to colour
bind_cols(default_test, pred = knn_5_pred) |> 
  arrange(default) |> 
  ggplot(aes(x = balance, y = income, colour = pred)) +
  geom_point(size = 1.3) + 
  scale_colour_viridis_d() +
  theme_minimal() +
  labs(title = "Predicted class (5nn)")

# there are quite some misclassifications, esp. many "No" predictions
# with "Yes" true class.
```
:::


::: {.callout-tip icon=false}
## 6. Repeat the same steps, but now with a `knn_2_pred` vector generated from a 2-nearest neighbours algorithm. Are there any differences?

```{r}
#| label: q6
#| include: !expr params$answers

knn_2_pred <- knn(
  train = default_train |> select(-default),
  test  = default_test  |> select(-default),
  cl    = as_factor(default_train$default),
  k     = 2
)

# second plot maps pred to colour
bind_cols(default_test, pred = knn_2_pred) |> 
  arrange(default) |> 
  ggplot(aes(x = balance, y = income, colour = pred)) +
  geom_point(size = 1.3) + 
  scale_colour_viridis_d() +
  theme_minimal() +
  labs(title = "Predicted class (2nn)")

# compared to the 5-nn model, more people get classified as "Yes"
# Still, the method is not perfect
```
:::


# Confusion matrix

The confusion matrix is an insightful summary of the plots we have made and the correct and incorrect classifications therein. A confusion matrix can be made in `R` with the `table()` function by entering two `factor` or `character` vectors:

```{r}
#| label: confmat

table(
  observed  = default_test$default, 
  predicted = knn_2_pred
)

```

::: {.callout-tip icon=false}
## 7. What would this confusion matrix look like if the classification were perfect?

```{r}
#| label: q7
#| include: !expr params$answers

# All the observations would fall in the yes-yes or no-no categories; 
# the off-diagonal elements would be 0 like so:

table(
  observed  = default_test$default, 
  predicted = default_test$default
)
```
:::


::: {.callout-tip icon=false}
## 8. Make a confusion matrix for the 5-nn model and compare it to that of the 2-nn model. What do you conclude?

```{r}
#| label: q8
#| include: !expr params$answers

table(
  observed  = default_test$default, 
  predicted = knn_5_pred
)

# the 2nn model has more true positives (yes-yes) but also more false
# positives (truly no but predicted yes). Overall the 5nn method has 
# slightly better accuracy (proportion of correct classifications).
```
:::

# Logistic regression

KNN directly predicts the class of a new observation using a majority vote of the existing observations closest to it. In contrast to this, logistic regression predicts the `log-odds` of belonging to category 1. These log-odds can then be transformed to probabilities by performing an inverse logit transform:

$$ p = \frac{1}{1+e^{-\alpha}}$$, where $\alpha$ indicates log-odds for being in class 1 and $p$ is the probability.

Therefore, logistic regression is a `probabilistic` classifier as opposed to a `direct` classifier such as KNN: indirectly, it outputs a probability which can then be used in conjunction with a cutoff (usually 0.5) to classify new observations.

Logistic regression in `R` happens with the `glm()` function, which stands for generalized linear model. Here we have to indicate that the residuals are modeled not as a gaussian (normal distribution), but as a `binomial` distribution.

::: {.callout-tip icon=false}
## 9. Use `glm()` with argument `family = binomial` to fit a logistic regression model `lr_mod` to the `default_train` data.

```{r}
#| label: q9
#| include: !expr params$answers

lr_mod <- glm(default ~ ., family = binomial, data = default_train)

```
:::

Now we have generated a model, we can use the `predict()` method to output the estimated probabilities for each point in the training dataset. By default `predict` outputs the log-odds, but we can transform it back using the inverse logit function of before or setting the argument `type = "response"` within the predict function. 

::: {.callout-tip icon=false}
## 10. Visualise the predicted probabilities versus observed class for the training dataset in `lr_mod`. 

You can choose for yourself which type of visualisation you would like to make. Write down your interpretations along with your plot.

```{r}
#| label: q10
#| include: !expr params$answers

tibble(observed  = default_train$default, 
       predicted = predict(lr_mod, type = "response")) |> 
  ggplot(aes(y = predicted, x = observed, colour = observed)) +
  geom_point(position = position_jitter(width = 0.2), alpha = .3) +
  scale_colour_manual(values = c("dark blue", "orange"), guide = "none") +
  theme_minimal() +
  labs(y = "Predicted probability to default")

# I opted for a raw data display of all the points in the test set. Here,
# we can see that the defaulting category has a higher average probability
# for a default compared to the "No" category, but there are still data 
# points in the "No" category with high predicted probability for defaulting.
```
:::


Another advantage of logistic regression is that we get coefficients we can interpret.


::: {.callout-tip icon=false}
## 11. Look at the coefficients of the `lr_mod` model and interpret the coefficient for `balance`. What would the probability of default be for a person who is not a student, has an income of 40000, and a balance of 3000 dollars at the end of each month? Is this what you expect based on the plots we've made before?

```{r}
#| label: q11
#| include: !expr params$answers

coefs <- coef(lr_mod)
coefs["balance"]
# The higher the balance, the higher the log-odds of defaulting. Precisely:
# Each dollar increase in balance increases the log-odds by 0.0058.

# Let's calculate the log-odds for our person
logodds <- coefs[1] + 4e4*coefs[4] + 3e3*coefs[3]

# Let's convert this to a probability
1 / (1 + exp(-logodds))

# probability of .998 of defaulting. This is in line with the plots of before
# because this new data point would be all the way on the right.
```
:::


## Visualising the effect of the balance variable

In two steps, we will visualise the effect `balance` has on the predicted default probability.

::: {.callout-tip icon=false}
## 12. Create a data frame called `balance_df` with 3 columns and 500 rows: `student` always 0, `balance` ranging from 0 to 3000, and `income` always the mean income in the `default_train` dataset.

```{r}
#| label: q12
#| include: !expr params$answers

balance_df <- tibble(
  student = rep(0, 500),
  balance = seq(0, 3000, length.out = 500),
  income  = rep(mean(default_train$income), 500)
)
```
:::

::: {.callout-tip icon=false}
## 13. Use this dataset as the `newdata` in a `predict()` call using `lr_mod` to output the predicted probabilities for different values of `balance`. Then create a plot with the `balance_df$balance` variable mapped to x and the predicted probabilities mapped to y. Is this in line with what you expect? 

```{r}
#| label: q13
#| include: !expr params$answers

balance_df$predprob <- predict(lr_mod, newdata = balance_df, type = "response")

balance_df |> 
  ggplot(aes(x = balance, y = predprob)) +
  geom_line(col = "dark blue", linewidth = 1) +
  theme_minimal()

# Just before 2000 in the first plot is where the ratio of
# defaults to non-defaults is 50-50. So this line is exactly what we expect!
```
:::


::: {.callout-tip icon=false}
## 14. Create a confusion matrix just as the one for the KNN models by using a cutoff predicted probability of 0.5. Does logistic regression perform better?

```{r}
#| label: q14
#| include: !expr params$answers

pred_prob <- predict(lr_mod, newdata = default_test, type = "response")
pred_lr   <- factor(pred_prob > .5, labels = c("No", "Yes"))

table(
  observed  = default_test$default, 
  predicted = pred_lr
)

# logistic regression performs better in every way than knn. This depends on
# your random split so your mileage may vary
```
:::


# Linear discriminant analysis

The last method we will use is LDA, using the `lda()` function from the `MASS` package.


::: {.callout-tip icon=false}
## 15. Train an LDA classifier `lda_mod` on the training set.

```{r}
#| label: q15
#| include: !expr params$answers

lda_mod <- lda(default ~ ., data = default_train)

```
:::

::: {.callout-tip icon=false}
## 16. Look at the `lda_mod` object. What can you conclude about the characteristics of the people who default on their loans?

```{r}
#| label: q16
#| include: !expr params$answers
lda_mod

# defaulters have a larger proportion of students than non-defaulters 
# (40% vs 29%), they have a slightly lower income, and they have a 
# much higher remaining credit card balance than non-defaulters.
```
:::


::: {.callout-tip icon=false}
## 17. Create a confusion matrix and compare it to the previous methods.

```{r}
#| label: q17
#| include: !expr params$answers

pred_lda <- predict(lda_mod, newdata = default_test)
table(
  observed  = default_test$default,
  predicted = pred_lda$class
)

# It has slightly more false negatives, a lower false positive rate, 
# and fewer true positives than logistic regression.
```
:::


# OPTIONAL: Final assignment

::: {.callout-tip icon=false}

## 18. Create a model (using knn, logistic regression, or LDA) to predict whether a 14 year old boy from the 3rd class would have survived the Titanic disaster. You can find the data [here](data/Titanic.csv). Would the passenger have survived if they were a girl in 2nd class?

```{r}
#| label: q18
#| include: !expr params$answers
#| message: false


titanic <- read_csv("data/Titanic.csv")

# I'll do a logistic regression with all interactions
lr_mod_titanic <- glm(Survived ~ PClass * Sex * Age, family = binomial, data = titanic)

predict(
  object = lr_mod_titanic, 
  newdata = tibble(
    PClass = c( "3rd",    "2nd"),
    Age    = c(    14,       14), 
    Sex    = c("male", "female")
  ), 
  type = "response"
)

# So our hypothetical passenger does not have a large survival probability:
# our model would classify the boy as not surviving. The girl would likely
# survive however. This is due to the women and children getting preferred
# access to the lifeboats. Also 3rd class was way below deck.
```
:::
